Towards a general mathematical theory of experimental science

### 2018-06-14 Articles/TowardsAGeneralMathematicalTheoryOfExperimentalScience/article.tex
Changed discussion

One should not confuse this approach with the idea of a grand-unified theory or theory of everything. In that program, one looks for the theory of fundamental constituents from which all other theories can be derived. Our program is altogether different and orthogonal. Our subject matter are the scientific models themselves. Our aim is to understand what aspects of a scientific theory are already constrained by the mere fact that it is a scientific theory or by the assumptions one bakes into them. While this general theory may be of some utility to someone working on a grand-unified theory as it may help narrow the search, neither one will lead to the other.

@@

To give an example, consider an experimental domain formed by statements like \statement{this quantity is between $a$ and $b$}. For example, \statement{the distance between the earth and the moon is between 384 and 385 thousand Km}. And suppose that the precision can be arbitrarily narrowed. We can define the following.

\begin{defn}
	An experimental domain $\edomain_X$ is said to allow \textbf{arbitrary precision} if it describes a quantity measurable with arbitrary precision. Formally, there exists a countable basis $\basis \subseteq \edomain_X$ such that:
	\begin{itemize}
		\item we can define a function $\operatorname{bounds} : \basis \to \mathbb{Q} \times \mathbb{Q}$ such that $\operatorname{bounds}(\stmt) = (a, b)$ is a tuple where $a < b$ for all $\stmt \in \basis$
		\item for every pair of rational numbers $(a, b) \in \mathbb{Q} \times \mathbb{Q}$ there exists an $\obs \in \basis$ such that $\operatorname{bounds}(\stmt) = (a, b)$
		\item if $\stmt = \stmt_1 \AND \stmt_2$, $\operatorname{bounds}(\stmt_1) = (a_1, b_1)$ and $\operatorname{bounds}(\stmt_2) = (a_2, b_2)$, then $\stmt \equiv \contradiction$ if $b_1 \leq a_2$ or $b_2 \leq a_1$ otherwise $\stmt \nequiv \contradiction$ and $\operatorname{bounds}(\stmt) = (\max(a_1, a_2), \min(b_1, b_2))$
	\end{itemize}
\end{defn}

Then it can be shown that:

\begin{thrm}[Arbitrary precision is continuity]\label{thrm_arbitrary_precision_is_continuity}
	Given an experimental domain $\edomain_X$, the set of possibilities with their natural topology is homeomorphic to the reals with the standard topology if and only if the domain allows arbitrary precision.
\end{thrm}

This can be proven by showing that each possibility is associated to an equivalence class of Cauchy sequences, which corresponds to a real number. Each open interval corresponds to a finite precision measurement. The possibilities are not themselves verifiable statements: we cannot measure a continuous quantity with infinite precision. But it is a theoretical statement: we can use the value to predict which finite precision measurements will be verified (the ones that contain that value).

This connection creates a strong bridge between the standard mathematical structures used in physics and their meaning. Every theorem, every proof on those structures can now be given a direct physical meaning as well. And given the foundational nature of topologies and $\sigma$-algebras, we hope to extend this framework to measure theory, differential geometry, symplectic geometry, Riemannian geometry, probability theory and so on.

>>>>

One should not confuse this approach with the idea of a grand-unified theory or theory of everything. In that program, one looks for the theory of fundamental constituents from which all other theories can be derived. Our program is altogether different and orthogonal. Our subject matter is the scientific models themselves. Our aim is to understand what aspects of a scientific theory are already constrained by the mere fact that it is a scientific theory or by the assumptions one bakes into them. While this general theory may be of some utility to someone working on a grand-unified theory as it may help narrow the search, neither one will lead to the other.

@@

\section{Brief discussion}

The connection we outlined creates a strong bridge between the standard mathematical structures used in physics and their meaning. Every theorem, every proof on those structures can now be given a direct physical meaning as well. And given the foundational nature of topologies and $\sigma$-algebras, we hope to extend this framework to measure theory, differential geometry, symplectic geometry, Riemannian geometry, probability theory\footnote{Probability statements will be of the form \statement{The position of the center of mass of the system is between 12 and 13 nm, 20\% to 21\% of the time.}} and so on.

To give an example, consider an experimental domain where the basis is composed of statements like \statement{this quantity is between $a$ and $b$} where $a$ and $b$ are different values that can be arbitrarily close. For example, \statement{the distance between the earth and the moon is between 384 and 385 thousand Km}. Each verifiable statement of the basis corresponds to an open interval of the real line, therefore we find a correspondence between arbitrary precision measurements and the standard topology on the reals, as this is the one generated by open intervals. Note, in fact, that this topology is second-countable and Kolmogorov.

The possibilities (i.e. \statement{the mass of the photon is precisely zero}) are not themselves verifiable since infinite precision measurements of a continuous quantity are not possible. Yet, their negation (i.e. \statement{the mass of the photon isn't precisely zero}) could be verified in practice. This is expressed mathematically by the fact that singletons are closed sets. The theoretical domain, instead, corresponds to the standard Borel algebra.

Similarly, one can imagine verifiable statements for relationships (i.e. \statement{when the temperature of the mercury column is between 24 and 25 C, its height is between 24 and 25 mm}) and for statistical variables (i.e. \statement{if this coin is tossed enough times, the ratio of heads will be between 45\% and 55\%}). Ultimately the general theory will need to show precisely what mathematical structures map to experimental domains formed by statements of these types and under what assumptions.
